---
title: "Getting Started with epiRomics"
author: "Alex M. Mawla & Mark O. Huising"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    df_print: default
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: yes
    df_print: default
    latex_engine: pdflatex
    keep_tex: no
vignette: |
  %\VignetteIndexEntry{Getting Started with epiRomics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE, width = 70)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE,
  cache    = FALSE,
  fig.width  = 12,
  fig.height = 8,
  out.width  = "100%",
  tidy       = FALSE
)

## ---- Locate example data ----
## Priority 1: installed package extdata (full install or R CMD check)
extdata_dir <- system.file("extdata", package = "epiRomics")
has_data <- nzchar(extdata_dir) &&
  file.exists(file.path(extdata_dir, "example_epiRomics_Db_sheet.csv")) &&
  dir.exists(file.path(extdata_dir, "BigWigs"))

## Priority 2: development source tree (rendering from package root)
if (!has_data) {
  dev_path <- normalizePath(file.path("..", "inst", "extdata"),
                            mustWork = FALSE)
  if (dir.exists(dev_path) &&
      file.exists(file.path(dev_path, "example_epiRomics_Db_sheet.csv")) &&
      dir.exists(file.path(dev_path, "BigWigs"))) {
    extdata_dir <- dev_path
    has_data <- TRUE
  }
}

## Priority 3: BiocFileCache (lightweight GitHub install + cached data)
if (!has_data) {
  has_data <- epiRomics::epiRomics_has_cache()
  if (has_data) extdata_dir <- epiRomics::epiRomics_cache_path()
}

## Priority 4: interactive download as last resort
if (!has_data && interactive()) {
  message("Downloading epiRomics example data (~1.3 GB, one-time)...")
  tryCatch(
    { epiRomics::epiRomics_cache_data(); has_data <- TRUE
      extdata_dir <- epiRomics::epiRomics_cache_path() },
    error = function(e) message("Download failed: ", e$message,
                                "\nRun epiRomics_cache_data() to retry.")
  )
}

if (!has_data) extdata_dir <- NULL

## BigWig plots require the signal files
has_bigwigs <- has_data && file.exists(
  file.path(extdata_dir, "BigWigs",
            "Kaestner_Alpha.atac.signal.subset.multi_chr.bigwig")
)
```

```{css, echo = FALSE}
pre code {
  white-space: pre-wrap;
  word-wrap: break-word;
  overflow-wrap: break-word;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
img {
  max-width: 100%;
  height: auto;
}
.note-box {
  background-color: #f0f7ff;
  border-left: 4px solid #3c78d8;
  padding: 12px 16px;
  margin: 16px 0;
  border-radius: 4px;
}
```

# Abstract

**Summary** epiRomics is an R package designed to integrate multi-omics
data in order to identify and visualize enhancer regions alongside gene
expression and other epigenomic modifications. Regulatory network
analysis can be done using combinatory approaches to infer regions of
significance such as enhancers, when combining ChIP and histone data.
Downstream analysis can identify co-occurrence of these regions of
interest with other user-supplied data, such as chromatin availability or
gene expression. Finally, this package allows for results to be
visualized at high resolution in a stand-alone browser.

**Availability and Implementation** epiRomics is released under the
Artistic-2.0 License. The source code and documents are freely available
through GitHub (<https://github.com/Huising-Lab/epiRomics>).

**Contact** ammawla@ucdavis.edu or mhuising@ucdavis.edu

**Supplementary information** Supplementary data and methods are
available online on *bioRxiv* or *GitHub*.

**Competing Interest Statement** The authors have declared no competing
interest.

# Citation

If you use epiRomics in published research, please cite:

Mawla AM, van der Meulen T, Huising MO. **Chromatin accessibility
differences between alpha, beta, and delta cells identifies common and
cell type-specific enhancers.** BMC Genomics. 2023;24:202.
<https://doi.org/10.1186/s12864-023-09293-6>

Mawla AM & Huising MO. **epiRomics: a multi-omics R package to identify
and visualize enhancers.** bioRxiv. 2021.
<https://doi.org/10.1101/2021.08.19.456732>

# Shiny epiRomics Showcase

Explore epiRomics results interactively through our companion Shiny web
application:

**[Shiny epiRomics](https://huisinglab.com/epiRomics_2021/index.html)** ---
Interactive browser for pancreatic islet enhancer and enhanceosome analysis
results.

# Example Data Setup {#data-setup}

::: {.note-box}
**Note:** epiRomics example data (~1.3 GB) is hosted remotely and
must be downloaded once before running this vignette with live output.
If the data is not yet cached, code chunks below will display the
source code without execution.

```r
## Run this once to download and cache example data
epiRomics_cache_data()
```
:::

```{r data-status, echo = FALSE, results = "asis"}
if (has_data) {
  cat("**Data status:** Example data is available.",
      "All code chunks will execute with live output.\n")
} else {
  cat(paste0(
    "**Data status:** Example data is **not** cached. ",
    "Code chunks are shown but not executed.\n\n",
    "Run `epiRomics::epiRomics_cache_data()` in an R session to ",
    "download the data, then rebuild this vignette.\n"
  ))
}
```

# Loading the epiRomics package

```{r load-library, eval = has_data}
library(epiRomics)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
```

# Brief explanation of example data

This package includes example data delineating human pancreatic islet
enhancers between alpha and beta cells.

Human pancreatic islet alpha and beta ATAC- and companion RNA-Seq data
were retrieved from GEO accession GSE76268 (Ackermann et al., 2016).

ATAC samples were processed using the ENCODE-DCC ATAC sequencing
pipeline, aligning to the hg38 build of the human genome (Harrow et
al., 2012; Consortium, 2012; Davis et al., 2018).

Peak calls generated through the pipeline using MACS2 (Zhang et al.,
2008) were analyzed downstream through the Bioconductor package DiffBind
(Ross-Innes et al., 2012) to identify differentially enriched chromatin
regions between the two cell types.

RNA samples were quality controlled using fastp (Chen et al., 2018), and
aligned using STAR (Dobin et al., 2013) to hg38. Wiggle files produced
by the STAR aligner were merged by cell type using UCSC command line
tools.

BigWig files were subsetted using UCSC command line tools (Kent et al.,
2010).

ChIP-seq peak calls for human pancreatic islet transcription factors
Foxa2, MafB, Nkx2.2, Nkx6.1, and Pdx1 were retrieved from EMBL-EBI
E-MTAB-1919 (Pasquali et al., 2014).

Histone modification peak calls for H3K27ac, H3K4me1, H3K27me3,
H3K9me3, H3K4me3, H3K36me3 were retrieved from GEO GSE16256 (Bernstein
et al., 2010), and H2A.Z from EMBL-EBI E-MTAB-1919 (Pasquali et al.,
2014).

All peak calls retrieved as hg19 were lifted over to hg38 using the
UCSC liftOver (Kent et al., 2002).

The FANTOM5 human enhancer database (Lizio et al., 2015), ultra-conserved
non-coding elements from the UCNE database (Dimitrieva and Bucher, 2012),
and the human islet regulome (Miguel-Escalada et al., 2019) were also
retrieved and lifted to hg38.

# Building the database

The database sheet CSV defines the input annotations. Each row specifies
a BED file with its name, relative path, genome build, format, and type.

Required columns: **name**, **path**, **genome**, **format**, **type**.
Genome must be in proper format (e.g., `mm10` or `hg38`).
Type can be: `histone`, `methyl`, `SNP`, `chip`, or `functional`.

```{r show-db-sheet, eval = has_data}
example_epiRomics_Db_sheet <- read.csv(
  file.path(extdata_dir,
    "example_epiRomics_Db_sheet.csv"))
head(example_epiRomics_Db_sheet, 5)
```

`epiRomics_build_dB()` constructs the S4 database from the data sheet.
The CSV uses relative paths; the `data_dir` parameter resolves them
against the cache directory automatically.

```{r build-db, eval = has_data, message = TRUE, warning = FALSE}
epiRomics_dB <- epiRomics_build_dB(
  epiRomics_db_file = file.path(extdata_dir,
    "example_epiRomics_Db_sheet.csv"),
  txdb_organism = paste0(
    "TxDb.Hsapiens.UCSC.hg38.knownGene::",
    "TxDb.Hsapiens.UCSC.hg38.knownGene"),
  epiRomics_genome = "hg38",
  epiRomics_organism = "org.Hs.eg.db",
  data_dir = extdata_dir)
```

# Quick visualization with epiRomics_quick_view

Before running the full enhancer pipeline, epiRomics provides a
zero-configuration entry point for the ultra-fast visualization of any gene locus
with BigWig signal tracks. No database setup is required---just pass a gene
symbol and BigWig file paths.

```{r setup-bigwigs, eval = has_data}
## Load BigWig track connection sheet and resolve paths
epiRomics_track_connection <- read.csv(
  file.path(extdata_dir,
    "example_epiRomics_BW_sheet.csv"))
epiRomics_track_connection$path <- file.path(
  extdata_dir,
  epiRomics_track_connection$path)

## Reorder: Beta first, then Alpha
beta_idx <- grep("Beta", epiRomics_track_connection$name)
alpha_idx <- grep("Alpha", epiRomics_track_connection$name)
epiRomics_track_connection <- epiRomics_track_connection[
  c(beta_idx, alpha_idx), ]

## Named vector of BigWig paths and colors
bw_paths <- setNames(
  epiRomics_track_connection$path,
  epiRomics_track_connection$name)
bw_colors <- epiRomics_track_connection$color
```

The `mirror = TRUE` option displays ATAC and RNA signals in a mirrored
layout for direct comparison between cell types.

## Beta cell genes

### INS

```{r fig-qv-ins, fig.width = 10, fig.height = 8, fig.align = "center", eval = has_bigwigs, warning = FALSE, message = FALSE}
epiRomics_quick_view("INS",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

### PDX1

```{r fig-qv-pdx1, fig.width = 10, fig.height = 8, fig.align = "center", eval = has_bigwigs, warning = FALSE, message = FALSE}
epiRomics_quick_view("PDX1",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

### MAFA

```{r fig-qv-mafa, fig.width = 10, fig.height = 8, fig.align = "center", eval = has_bigwigs, warning = FALSE, message = FALSE}
epiRomics_quick_view("MAFA",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

## Alpha cell genes

### GCG

```{r fig-qv-gcg, fig.width = 10, fig.height = 8, fig.align = "center", eval = has_bigwigs, warning = FALSE, message = FALSE}
epiRomics_quick_view("GCG",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```



### MAFB

```{r fig-qv-mafb, fig.width = 10, fig.height = 8, fig.align = "center", eval = has_bigwigs, warning = FALSE, message = FALSE}
epiRomics_quick_view("MAFB",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

## Both Alpha and Beta Cell Genes

### UCN3

```{r fig-qv-ucn3, fig.width = 10, fig.height = 8, fig.align = "center", eval = has_bigwigs, warning = FALSE, message = FALSE}
epiRomics_quick_view("UCN3",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

# Identifying putative enhancers

There is a lot of flexibility for data exploration here. In this
example, we search for putative enhancers using two histone marks known
to co-occur at enhancer regions: H3K4me1 and H3K27ac.

```{r putative-enhancers, eval = has_data}
epiRomics_putative_enhancers <-
  epiRomics_enhancers_co_marks(
    epiRomics_dB,
    epiRomics_histone_mark_1 = "h3k4me1",
    epiRomics_histone_mark_2 = "h3k27ac")

## First 5 putative enhancers
head(as.data.frame(
  epiRomics_putative_enhancers@annotations), 5)

```

# Cross-referencing enhancer calls to databases

## FANTOM Enhancer Database

Cross-referencing putative enhancers against the FANTOM5 enhancer
database increases confidence in the calls.

```{r filter-fantom, eval = has_data}
epiRomics_putative_enhancers_filtered_fantom <-
  epiRomics_enhancers_filter(
    epiRomics_putative_enhancers,
    epiRomics_dB,
    epiRomics_type = "hg38_custom_fantom")

head(as.data.frame(
  epiRomics_putative_enhancers_filtered_fantom@annotations), 5)
```

## Human Islet Regulome Active Enhancers

```{r filter-regulome-active, eval = has_data}
epiRomics_putative_enhancers_filtered_regulome_active <-
  epiRomics_enhancers_filter(
    epiRomics_putative_enhancers,
    epiRomics_dB,
    epiRomics_type = "hg38_custom_regulome_active")

head(as.data.frame(
  epiRomics_putative_enhancers_filtered_regulome_active@annotations), 5)
```

## Human Islet Regulome Super Enhancers

```{r filter-regulome-super, eval = has_data}
epiRomics_putative_enhancers_filtered_regulome_super <-
  epiRomics_enhancers_filter(
    epiRomics_putative_enhancers,
    epiRomics_dB,
    epiRomics_type = "hg38_custom_regulome_super")

head(as.data.frame(
  epiRomics_putative_enhancers_filtered_regulome_super@annotations), 5)
```

## Ultra-Conserved Non-Coding Elements (UCNEs)

```{r filter-ucnes, eval = has_data}
epiRomics_putative_enhancers_filtered_ucnes <-
  epiRomics_enhancers_filter(
    epiRomics_putative_enhancers,
    epiRomics_dB,
    epiRomics_type = "hg38_custom_ucnes")

head(as.data.frame(
  epiRomics_putative_enhancers_filtered_ucnes@annotations), 5)
```

# Screening for high TF co-binding sites (enhanceosomes)

Enhancers can be redundant and not all play an active regulatory role.
By intersecting putative enhancer calls with all available ChIP-seq data,
we identify enhanceosome regions with high transcription factor
co-binding. Results are sorted by co-binding count.

```{r enhanceosome-fantom, eval = has_data}
epiRomics_putative_enhanceosome_fantom <-
  epiRomics_enhanceosome(
    epiRomics_putative_enhancers_filtered_fantom,
    epiRomics_dB)

head(as.data.frame(
  epiRomics_putative_enhanceosome_fantom@annotations), 5)

## Evaluate calls on chromosome 1
fantom_df <- as.data.frame(
  epiRomics_putative_enhanceosome_fantom@annotations)
head(fantom_df[fantom_df$seqnames == "chr1", ], 5)

## Total enhanceosome regions
length(epiRomics_putative_enhanceosome_fantom@annotations)
```

# Transcription factor co-binding analysis

`epiRomics_tf_cobinding()` uses Fisher's exact test to identify
statistically significant TF co-binding pairs within enhanceosome
regions, reporting odds ratios and pointwise mutual information (PMI).

```{r cobinding, eval = has_data}
tf_cobinding <- epiRomics_tf_cobinding(
  epiRomics_putative_enhanceosome_fantom,
  epiRomics_dB)

tf_cobinding$pairwise
```

# Visualizing enhanceosome regions with signal tracks

```{r fig-fantom-track, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer(
  epiRomics_putative_enhanceosome_fantom,
  epiRomics_index = which(names(
    epiRomics_putative_enhanceosome_fantom@annotations) == "14"),
  epiRomics_dB = epiRomics_dB,
  epiRomics_track_connection = epiRomics_track_connection)
```

## Regulome active-filtered enhanceosomes

```{r enhanceosome-regulome-active, eval = has_data}
epiRomics_putative_enhanceosome_regulome_active <-
  epiRomics_enhanceosome(
    epiRomics_putative_enhancers_filtered_regulome_active,
    epiRomics_dB)

head(as.data.frame(
  epiRomics_putative_enhanceosome_regulome_active@annotations), 5)

reg_active_df <- as.data.frame(
  epiRomics_putative_enhanceosome_regulome_active@annotations)
head(reg_active_df[reg_active_df$seqnames == "chr1", ], 5)

length(
  epiRomics_putative_enhanceosome_regulome_active@annotations)
```

```{r fig-regulome-active-track, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer(
  epiRomics_putative_enhanceosome_regulome_active,
  epiRomics_index = which(names(
    epiRomics_putative_enhanceosome_regulome_active@annotations) == "21"),
  epiRomics_dB = epiRomics_dB,
  epiRomics_track_connection = epiRomics_track_connection)
```

## Regulome super-enhancer filtered enhanceosomes

```{r enhanceosome-regulome-super, eval = has_data}
epiRomics_putative_enhanceosome_regulome_super <-
  epiRomics_enhanceosome(
    epiRomics_putative_enhancers_filtered_regulome_super,
    epiRomics_dB)

head(as.data.frame(
  epiRomics_putative_enhanceosome_regulome_super@annotations), 5)

reg_super_df <- as.data.frame(
  epiRomics_putative_enhanceosome_regulome_super@annotations)
head(reg_super_df[reg_super_df$seqnames == "chr1", ], 5)

length(
  epiRomics_putative_enhanceosome_regulome_super@annotations)
```

```{r fig-regulome-super-track, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer(
  epiRomics_putative_enhanceosome_regulome_super,
  epiRomics_index = which(names(
    epiRomics_putative_enhanceosome_regulome_super@annotations) == "13"),
  epiRomics_dB = epiRomics_dB,
  epiRomics_track_connection = epiRomics_track_connection)
```

## UCNE-filtered enhanceosomes

```{r enhanceosome-ucnes, eval = has_data}
epiRomics_putative_enhanceosome_ucnes <-
  epiRomics_enhanceosome(
    epiRomics_putative_enhancers_filtered_ucnes,
    epiRomics_dB)

head(as.data.frame(
  epiRomics_putative_enhanceosome_ucnes@annotations), 5)

## Evaluate calls on chromosome 1
ucne_df <- as.data.frame(
  epiRomics_putative_enhanceosome_ucnes@annotations)
head(ucne_df[ucne_df$seqnames == "chr1", ], 5)

## Total enhanceosome regions
length(epiRomics_putative_enhanceosome_ucnes@annotations)
```

```{r fig-ucne-track, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer(
  epiRomics_putative_enhanceosome_ucnes,
  epiRomics_index = 4,
  epiRomics_dB = epiRomics_dB,
  epiRomics_track_connection = epiRomics_track_connection)
```


# Integrating differential chromatin accessibility

How can we use putative enhanceosome regions to infer biology between
cell states? Here we integrate DiffBind ATAC-seq data showing
differential chromatin accessibility between alpha and beta cells.

```{r integrate-chromatin, eval = has_data}
b.v.a <- read.csv(
  file.path(extdata_dir,
    "DBA_Beta_Versus_Alpha.csv"))
b.v.a <- GRanges(b.v.a)

## Filter for beta-enriched chromatin regions
beta.enriched <- b.v.a[b.v.a$Fold >= 1 & b.v.a$Conc>=3, ]

## Connect to putative enhanceosomes
beta_enhancer_regions <-
  epiRomics_regions_of_interest(
    epiRomics_putative_enhanceosome_fantom,
    beta.enriched)

head(as.data.frame(
  beta_enhancer_regions@annotations), 5)
```

```{r fig-beta-track, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer(
  beta_enhancer_regions,
  epiRomics_index = which(names(
    beta_enhancer_regions@annotations) == "305"),
  epiRomics_dB = epiRomics_dB,
  epiRomics_track_connection = epiRomics_track_connection)
```

# Gene-centered visualization

`epiRomics_track_layer_gene()` provides a gene-focused view combining
all database tracks (histone marks, ChIP, annotations) with BigWig
signal---no enhanceosome pipeline required.

```{r fig-gene-ins, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer_gene("INS",
  epiRomics_dB,
  epiRomics_track_connection)
```

```{r fig-gene-gcg, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer_gene("GCG",
  epiRomics_dB,
  epiRomics_track_connection)
```

```{r fig-gene-pdx1, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
## Wider padding captures distal regulatory elements
epiRomics_track_layer_gene("PDX1",
  epiRomics_dB,
  epiRomics_track_connection,
  padding = 5000)
```

# Chromatin state classification

`epiRomics_chromatin_states()` classifies genomic regions by their
histone mark combinations into six biologically meaningful states:
active, bivalent, poised, primed, repressed, and unmarked. Classification
follows ChromHMM/Roadmap Epigenomics conventions (Ernst & Kellis 2012;
Kundaje et al. 2015), with TSS proximity refinement.


```{r chromatin-states, eval = has_data}
chromatin_states <-
  epiRomics_chromatin_states(epiRomics_dB)

table(chromatin_states$chromatin_state)
table(chromatin_states$genomic_context)


pdx1_states <- chromatin_states[
  chromatin_states$seqnames == "chr13" &
    chromatin_states$start >= 27900000 &
    chromatin_states$end   <= 28000000, ]
head(pdx1_states, 10)
```

### Visualize the PDX1 locus with chromatin state tracks overlaid:

```{r fig-chromatin-pdx1, fig.width = 14, fig.height = 12, fig.align = "center", eval = has_bigwigs, message=FALSE}
epiRomics_track_layer_gene("PDX1",
  epiRomics_dB,
  epiRomics_track_connection,
  chromatin_states = chromatin_states,
  show_chromatin = TRUE,
  show_enhancer_highlight = TRUE,
  padding = 5000)
```


# Session Information

```{r sessioninfo}
sessionInfo()
```
