---
title: "Full Analysis with epiRomics"
author: "Alex M. Mawla & Mark O. Huising"
date: "`r Sys.Date()`"
bibliography: ../references.bib
nocite: |
  @mawla2023, @mawla2021
output:
  html_document:
    toc: yes
    toc_float: yes
    df_print: default
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: yes
    df_print: default
    latex_engine: pdflatex
    keep_tex: no
vignette: |
  %\VignetteIndexEntry{Full Analysis with epiRomics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

<!--
  Full-scale walkthrough backed by the 1.3 GB Zenodo example archive
  (cached via `epiRomics::cache_data()`). New users should start with
  the lightweight "Getting Started with epiRomics" vignette, which uses
  a small toy subset and requires no download.
-->


```{r knitr-setup, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE, width = 70)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = TRUE,
  cache     = FALSE,
  fig.width  = 12,
  fig.height = 8,
  fig.align  = "center",
  message    = FALSE,
  warning    = FALSE,
  out.width  = "100%",
  tidy       = FALSE
)
```

```{css, echo = FALSE}
pre code {
  white-space: pre-wrap;
  word-wrap: break-word;
  overflow-wrap: break-word;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
img {
  max-width: 100%;
  height: auto;
}
.note-box {
  background-color: #f0f7ff;
  border-left: 4px solid #3c78d8;
  padding: 12px 16px;
  margin: 16px 0;
  border-radius: 4px;
}
```

# Abstract

**Summary** epiRomics is an R package designed to integrate multi-omics
data in order to identify and visualize enhancer regions alongside gene
expression and other epigenomic modifications. Regulatory network
analysis can be done using combinatory approaches to infer regions of
significance such as enhancers, when combining ChIP and histone data.
Downstream analysis can identify co-occurrence of these regions of
interest with other user-supplied data, such as chromatin availability or
gene expression. Finally, this package allows for results to be
visualized at high resolution in a stand-alone browser.

**Availability and Implementation** epiRomics is released under the
Artistic-2.0 License. The source code and documents are freely available
through GitHub (<https://github.com/Huising-Lab/epiRomics>).

**Contact** ammawla@ucdavis.edu or mhuising@ucdavis.edu

**Supplementary information** Supplementary data and methods are
available online on *bioRxiv* or *GitHub*.

**Competing Interest Statement** The authors have declared no competing
interest.

# Citation

If you use epiRomics in published research, please cite:

Mawla AM, van der Meulen T, Huising MO. **Chromatin accessibility
differences between alpha, beta, and delta cells identifies common and
cell type-specific enhancers.** BMC Genomics. 2023;24:202.
<https://doi.org/10.1186/s12864-023-09293-6>

Mawla AM & Huising MO. **epiRomics: a multi-omics R package to identify
and visualize enhancers.** bioRxiv. 2021.
<https://doi.org/10.1101/2021.08.19.456732>

# Interactive showcases

Two companion Shiny applications render published enhanceosome
browsers built with epiRomics:

**[Shiny epiRomics Mouse Islet Enhanceosome Browser](https://www.huisinglab.com/epiRomics_2021/index.html)** ---
mouse pancreatic alpha, beta, and delta cell results from
[Mawla et al. 2023](https://doi.org/10.1186/s12864-023-09293-6).

**[Shiny epiRomics Human Islet Enhanceosome Browser](https://huisinglab.shinyapps.io/Server_epiRomics_Shiny_Human/)** ---
human alpha and beta cell results used as the example dataset for
this package (Mawla and Huising 2021).

# Installing and loading the epiRomics package

Install `epiRomics` and its annotation dependencies from
Bioconductor (these are declared as `Suggests`, so install them
explicitly the first time you render the vignette):

```r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c(
    "epiRomics",
    "TxDb.Hsapiens.UCSC.hg38.knownGene",
    "org.Hs.eg.db"
))
```

With the package and its annotation companions installed, attach
them at the top of the session:

```{r load-library}
library(epiRomics)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
```

`epiRomics` provides the analysis API; the hg38 `TxDb` supplies
gene models and the `org.Hs.eg.db` package resolves gene symbols.

# Locating the example data {#data-setup}

This vignette runs end-to-end against the full 1.3 GB curated
archive on Zenodo (see the Zenodo record at
<https://zenodo.org/records/19189987> and the package README for
curation methodology). Because the archive is too large to ship
inside a Bioconductor tarball, we download it once via
`cache_data()` and then resolve its on-disk location at render
time.

::: {.note-box}
**Run this once** (interactive session) to populate the
`BiocFileCache` entry used by every chunk below:

```r
epiRomics::cache_data()
```
:::

The block below probes four sources in priority order — the
installed package `extdata/`, the development source tree, the
on-disk `BiocFileCache`, and (only in an interactive session) an
automatic `cache_data()` fallback — then reports which source
was picked and whether the BigWig signal tracks are present:

```{r locate-data}
## Priority 1: installed package extdata (full install or R CMD check)
extdata_dir <- system.file("extdata", package = "epiRomics")
has_data <- nzchar(extdata_dir) &&
  file.exists(file.path(extdata_dir, "example_epiRomics_Db_sheet.csv")) &&
  dir.exists(file.path(extdata_dir, "BigWigs"))

## Priority 2: development source tree (rendering from package root)
if (!has_data) {
  dev_path <- normalizePath(file.path("..", "inst", "extdata"),
                            mustWork = FALSE)
  if (dir.exists(dev_path) &&
      file.exists(file.path(dev_path,
                            "example_epiRomics_Db_sheet.csv")) &&
      dir.exists(file.path(dev_path, "BigWigs"))) {
    extdata_dir <- dev_path
    has_data <- TRUE
  }
}

## Priority 3: BiocFileCache (lightweight install + cached data)
if (!has_data && isTRUE(epiRomics::has_cache())) {
  cache_dir <- epiRomics::get_cache_path()
  if (!is.null(cache_dir) &&
      file.exists(file.path(cache_dir,
                            "example_epiRomics_Db_sheet.csv")) &&
      dir.exists(file.path(cache_dir, "BigWigs"))) {
    extdata_dir <- cache_dir
    has_data <- TRUE
  }
}

## Priority 4: interactive download as last resort
if (!has_data && interactive()) {
  message("Downloading epiRomics example data (~1.3 GB, one-time)...")
  tryCatch(
    { epiRomics::cache_data(); has_data <- TRUE
      extdata_dir <- epiRomics::get_cache_path() },
    error = function(e) message("Download failed: ", e$message,
                                "\nRun cache_data() to retry.")
  )
}

if (!has_data) extdata_dir <- NULL

## BigWig plots require the signal files
has_bigwigs <- has_data && file.exists(
  file.path(extdata_dir, "BigWigs",
            "Kaestner_Alpha.atac.signal.subset.multi_chr.bigwig")
)

## Report a machine-independent discovery descriptor rather than the
## absolute path, so the rendered vignette is identical across users.
data_source <- if (!has_data) {
  "none (run epiRomics::cache_data() in an interactive session)"
} else if (identical(extdata_dir,
                     system.file("extdata", package = "epiRomics"))) {
  "installed package extdata/"
} else if (grepl("inst/extdata$", extdata_dir)) {
  "development source tree (inst/extdata/)"
} else {
  "BiocFileCache (epiRomics::cache_data())"
}

has_data
has_bigwigs
data_source
```

`has_data` and `has_bigwigs` drive the per-chunk `eval =` guards
below: chunks that read BED/CSV files are gated on `has_data`;
chunks that plot BigWig signal are gated on `has_bigwigs`. If the
archive is not yet cached, the vignette still renders — the gated
chunks simply skip — but running `epiRomics::cache_data()` once
in an interactive session unlocks the full walkthrough.

```{r data-status, echo = FALSE, results = "asis"}
if (has_data) {
  cat("**Data status:** Example data is available.",
      "All code chunks will execute with live output.\n")
} else {
  cat(paste0(
    "**Data status:** Example data is **not** cached. ",
    "Code chunks are shown but not executed.\n\n",
    "Run `epiRomics::cache_data()` in an R session to ",
    "download the data, then rebuild this vignette.\n"
  ))
}
```

# Brief explanation of example data

This package includes example data delineating human pancreatic islet
enhancers between alpha and beta cells.

Human pancreatic islet alpha and beta ATAC- and companion RNA-Seq data
were retrieved from GEO accession GSE76268 [@ackermann2016].

ATAC samples were processed using the ENCODE-DCC ATAC sequencing
pipeline, aligning to the hg38 build of the human genome
[@harrow2012; @encode2012; @davis2018].

Peak calls generated through the pipeline using MACS2 [@zhang2008] were
analyzed downstream through the Bioconductor package DiffBind
[@ross_innes2012] to identify differentially enriched chromatin
regions between the two cell types.

RNA samples were quality controlled using fastp [@chen2018], and
aligned using STAR [@dobin2013] to hg38. Wiggle files produced
by the STAR aligner were merged by cell type using UCSC command line
tools.

BigWig files were subsetted using UCSC command line tools [@kent2010].

ChIP-seq peak calls for human pancreatic islet transcription factors
Foxa2, MafB, Nkx2.2, Nkx6.1, and Pdx1 were retrieved from EMBL-EBI
E-MTAB-1919 [@pasquali2014].

Histone modification peak calls for H3K27ac, H3K4me1, H3K27me3,
H3K9me3, H3K4me3, H3K36me3 were retrieved from GEO GSE16256
[@bernstein2010], and H2A.Z from EMBL-EBI E-MTAB-1919 [@pasquali2014].

All peak calls retrieved as hg19 were lifted over to hg38 using the
UCSC liftOver [@kent2002].

The FANTOM5 human enhancer database [@lizio2015], ultra-conserved
non-coding elements from the UCNE database [@dimitrieva2012], and the
human islet regulome [@miguel_escalada2019] were also retrieved and
lifted to hg38.

# Building the database

The database sheet CSV defines the input annotations. Each row specifies
a BED file with its name, relative path, genome build, format, and type.

Required columns: **name**, **path**, **genome**, **format**, **type**.
`genome` is a free-form string (`hg38`, `mm10`, `rn6`, …) that is
cross-checked against the TxDb you supply to `build_database()` via
`GenomeInfoDb::genome(txdb)`, so a mismatch raises an error rather than
proceeding silently. Any organism available in Bioconductor's annotation
catalogue works; the code path does not branch on a whitelist.
Type can be: `histone`, `methyl`, `SNP`, `chip`, or `functional`.

```{r show-db-sheet, eval = has_data}
example_epiRomics_Db_sheet <- read.csv(
  file.path(extdata_dir,
    "example_epiRomics_Db_sheet.csv"))
head(example_epiRomics_Db_sheet, 5)
```

`build_database()` constructs the S4 database from the data sheet.
The CSV uses relative paths; the `data_dir` parameter resolves them
against the cache directory automatically.

```{r build-db, eval = has_data, message = TRUE, warning = FALSE}
database <- build_database(
  db_file = file.path(extdata_dir,
    "example_epiRomics_Db_sheet.csv"),
  txdb_organism = paste0(
    "TxDb.Hsapiens.UCSC.hg38.knownGene::",
    "TxDb.Hsapiens.UCSC.hg38.knownGene"),
  genome = "hg38",
  organism = "org.Hs.eg.db",
  data_dir = extdata_dir)
```

`database` is the S4 `epiRomicsS4` object used by every downstream
step. It holds the overlap-annotated `GRanges` and metadata; all
pipeline functions accept it directly.

# Quick visualization with plot_quick_view

Before running the full enhancer pipeline, epiRomics provides a
zero-configuration entry point for the ultra-fast visualization of any gene
locus with BigWig signal tracks. No database setup is required---just pass
a gene symbol and BigWig file paths.

```{r setup-bigwigs, eval = has_data}
## Load BigWig track connection sheet and resolve paths
track_connection <- read.csv(
  file.path(extdata_dir,
    "example_epiRomics_BW_sheet.csv"))
track_connection$path <- file.path(
  extdata_dir,
  track_connection$path)

## Reorder: Beta first, then Alpha
beta_idx <- grep("Beta", track_connection$name)
alpha_idx <- grep("Alpha", track_connection$name)
track_connection <- track_connection[
  c(beta_idx, alpha_idx), ]

## Named vector of BigWig paths and colors
bw_paths <- setNames(
  track_connection$path,
  track_connection$name)
bw_colors <- track_connection$color
```

The `mirror = TRUE` option displays ATAC and RNA signals in a mirrored
layout for direct comparison between cell types. The `genome = "hg38"`
argument below is a string label that `plot_quick_view()` uses to pick
the matching `TxDb.*` / `org.*.db` pair loaded in the session; any
genome label that corresponds to an installed Bioconductor
`TxDb.*` package (`mm10`, `rn6`, `dm6`, …) works identically.

## Beta cell genes

### INS

The *INS* locus encodes insulin and is the canonical beta-cell marker.
We expect sharp beta-specific ATAC accessibility and RNA signal here;
mirrored against the alpha-cell tracks, the asymmetry should be visible
at a glance.

```{r fig-qv-ins, fig.width = 10, fig.height = 8, eval = has_bigwigs}
plot_quick_view("INS",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

The beta-cell ATAC and RNA tracks (above the axis) clearly dominate
the *INS* body while alpha-cell signal (below the axis) is nearly
flat, confirming the expected cell-type-specific activity.

### PDX1

*PDX1* is the beta-cell master transcription factor. Unlike *INS*,
much of its regulatory activity sits in distal enhancers rather than
directly over the gene body, so we expect strong beta-specific
accessibility spread across a broader locus.

```{r fig-qv-pdx1, fig.width = 10, fig.height = 8, eval = has_bigwigs}
plot_quick_view("PDX1",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

The beta-cell ATAC signal spans the gene body and its flanking
regulatory region; RNA expression is beta-biased but less extreme
than *INS* because *PDX1* is expressed at lower steady-state levels.

### MAFA

*MAFA* is a late-stage beta-cell identity factor.

```{r fig-qv-mafa, fig.width = 10, fig.height = 8, eval = has_bigwigs}
plot_quick_view("MAFA",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

As with *INS* and *PDX1*, the beta-cell tracks dominate; alpha-cell
signal is sparse.

## Alpha cell genes

### GCG

*GCG* encodes glucagon, the alpha-cell counterpart to insulin. The
expected pattern here is a mirror image of *INS*: strong alpha-cell
ATAC and RNA signal, minimal beta-cell activity.

```{r fig-qv-gcg, fig.width = 10, fig.height = 8, eval = has_bigwigs}
plot_quick_view("GCG",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

The alpha-cell tracks now dominate the gene body; beta-cell tracks
are close to baseline, exactly mirroring the *INS* pattern above.

### MAFB

*MAFB* is the alpha-cell identity factor (mouse *Mafa*-like role).

```{r fig-qv-mafb, fig.width = 10, fig.height = 8, eval = has_bigwigs}
plot_quick_view("MAFB",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

The alpha-cell ATAC and RNA tracks show the expected alpha-biased
signal; MAFB is expressed in both cell types developmentally but
remains more abundant in adult alpha cells.

## Both Alpha and Beta Cell Genes

### UCN3

*UCN3* (urocortin-3) is expressed in both alpha and beta cells,
though most strongly in mature beta cells. We therefore expect
some signal in both tracks, with a beta-cell bias.

```{r fig-qv-ucn3, fig.width = 10, fig.height = 8, eval = has_bigwigs}
plot_quick_view("UCN3",
  bw_paths = bw_paths,
  colors = bw_colors,
  mirror = TRUE,
  genome = "hg38")
```

Both cell types show signal over the gene body, with the beta-cell
side typically stronger — a useful contrast to the strictly
cell-type-specific *INS* and *GCG* loci above.

# Identifying putative enhancers

There is a lot of flexibility for data exploration here. In this
example, we search for putative enhancers using two histone marks known
to co-occur at enhancer regions: H3K4me1 and H3K27ac.

```{r putative-enhancers, eval = has_data}
putative_enhancers <-
  find_enhancers_by_comarks(
    database,
    histone_mark_1 = "h3k4me1",
    histone_mark_2 = "h3k27ac")

## First 5 putative enhancers
head(as.data.frame(
  methods::slot(putative_enhancers, "annotations")), 5)

```

# Cross-referencing enhancer calls to databases

## FANTOM Enhancer Database

Cross-referencing putative enhancers against the FANTOM5 enhancer
database increases confidence in the calls.

```{r filter-fantom, eval = has_data}
putative_enhancers_filtered_fantom <-
  filter_enhancers(
    putative_enhancers,
    database,
    type = "hg38_custom_fantom")

head(as.data.frame(
  methods::slot(putative_enhancers_filtered_fantom, "annotations")), 5)
```

## Human Islet Regulome Active Enhancers

```{r filter-regulome-active, eval = has_data}
putative_enhancers_filtered_regulome_active <-
  filter_enhancers(
    putative_enhancers,
    database,
    type = "hg38_custom_regulome_active")

head(as.data.frame(
  methods::slot(putative_enhancers_filtered_regulome_active, "annotations")), 5)
```

## Human Islet Regulome Super Enhancers

```{r filter-regulome-super, eval = has_data}
putative_enhancers_filtered_regulome_super <-
  filter_enhancers(
    putative_enhancers,
    database,
    type = "hg38_custom_regulome_super")

head(as.data.frame(
  methods::slot(putative_enhancers_filtered_regulome_super, "annotations")), 5)
```

## Ultra-Conserved Non-Coding Elements (UCNEs)

```{r filter-ucnes, eval = has_data}
putative_enhancers_filtered_ucnes <-
  filter_enhancers(
    putative_enhancers,
    database,
    type = "hg38_custom_ucnes")

head(as.data.frame(
  methods::slot(putative_enhancers_filtered_ucnes, "annotations")), 5)
```

# Screening for high TF co-binding sites (enhanceosomes)

Enhancers can be redundant and not all play an active regulatory role.
By intersecting putative enhancer calls with all available ChIP-seq data,
we identify enhanceosome regions with high transcription factor
co-binding. Results are sorted by co-binding count.

```{r enhanceosome-fantom, eval = has_data}
putative_enhanceosome_fantom <-
  find_enhanceosomes(
    putative_enhancers_filtered_fantom,
    database)

head(as.data.frame(
  methods::slot(putative_enhanceosome_fantom, "annotations")), 5)

## Evaluate calls on chromosome 1
fantom_df <- as.data.frame(
  methods::slot(putative_enhanceosome_fantom, "annotations"))
head(fantom_df[fantom_df$seqnames == "chr1", ], 5)

## Total enhanceosome regions
length(methods::slot(putative_enhanceosome_fantom, "annotations"))
```

# Transcription factor co-binding analysis

`analyze_tf_cobinding()` uses Fisher's exact test to identify
statistically significant TF co-binding pairs within enhanceosome
regions, reporting odds ratios and pointwise mutual information (PMI).

```{r cobinding, eval = has_data}
tf_cobinding <- analyze_tf_cobinding(
  putative_enhanceosome_fantom,
  database)

tf_cobinding$pairwise
```

# Visualizing enhanceosome regions with signal tracks

```{r fig-fantom-track, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_tracks(
  putative_enhanceosome_fantom,
  index = which(names(
    methods::slot(putative_enhanceosome_fantom, "annotations")) == "14"),
  database = database,
  track_connection = track_connection)
```

## Regulome active-filtered enhanceosomes

```{r enhanceosome-regulome-active, eval = has_data}
putative_enhanceosome_regulome_active <-
  find_enhanceosomes(
    putative_enhancers_filtered_regulome_active,
    database)

head(as.data.frame(
  methods::slot(putative_enhanceosome_regulome_active, "annotations")), 5)

reg_active_df <- as.data.frame(
  methods::slot(putative_enhanceosome_regulome_active, "annotations"))
head(reg_active_df[reg_active_df$seqnames == "chr1", ], 5)

length(
  methods::slot(putative_enhanceosome_regulome_active, "annotations"))
```

The head of the data frame shows the coordinates and per-TF hit
counts of the first five enhanceosomes that overlap active Islet
Regulome elements, and the final `length()` prints the total number
of such regions. We pick one for track visualization below.

```{r fig-reg-active-track, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_tracks(
  putative_enhanceosome_regulome_active,
  index = which(
    names(methods::slot(
      putative_enhanceosome_regulome_active,
      "annotations")) == "21"),
  database = database,
  track_connection = track_connection)
```

## Regulome super-enhancer filtered enhanceosomes

```{r enhanceosome-regulome-super, eval = has_data}
putative_enhanceosome_regulome_super <-
  find_enhanceosomes(
    putative_enhancers_filtered_regulome_super,
    database)

head(as.data.frame(
  methods::slot(putative_enhanceosome_regulome_super, "annotations")), 5)

reg_super_df <- as.data.frame(
  methods::slot(putative_enhanceosome_regulome_super, "annotations"))
head(reg_super_df[reg_super_df$seqnames == "chr1", ], 5)

length(
  methods::slot(putative_enhanceosome_regulome_super, "annotations"))
```

As before, the printed data frame and total count summarise the
enhanceosome set narrowed to super-enhancer calls; the track plot
below renders a representative locus.

```{r fig-reg-super-track, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_tracks(
  putative_enhanceosome_regulome_super,
  index = which(
    names(methods::slot(
      putative_enhanceosome_regulome_super,
      "annotations")) == "13"),
  database = database,
  track_connection = track_connection)
```

## UCNE-filtered enhanceosomes

```{r enhanceosome-ucnes, eval = has_data}
putative_enhanceosome_ucnes <-
  find_enhanceosomes(
    putative_enhancers_filtered_ucnes,
    database)

head(as.data.frame(
  methods::slot(putative_enhanceosome_ucnes, "annotations")), 5)

## Evaluate calls on chromosome 1
ucne_df <- as.data.frame(
  methods::slot(putative_enhanceosome_ucnes, "annotations"))
head(ucne_df[ucne_df$seqnames == "chr1", ], 5)

## Total enhanceosome regions
length(methods::slot(putative_enhanceosome_ucnes, "annotations"))
```

The UCNE-filtered result isolates enhanceosomes that sit inside
ultra-conserved non-coding elements — a small, high-confidence
subset. The track plot below visualises one of these regions.

```{r fig-ucne-track, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_tracks(
  putative_enhanceosome_ucnes,
  index = 4,
  database = database,
  track_connection = track_connection)
```


# Integrating differential chromatin accessibility

How can we use putative enhanceosome regions to infer biology between
cell states? Here we integrate DiffBind ATAC-seq data showing
differential chromatin accessibility between alpha and beta cells.

```{r integrate-chromatin, eval = has_data}
b.v.a <- read.csv(
  file.path(extdata_dir,
    "DBA_Beta_Versus_Alpha.csv"))
b.v.a <- GRanges(b.v.a)

## Filter for beta-enriched chromatin regions
beta.enriched <- b.v.a[b.v.a$Fold >= 1 & b.v.a$Conc>=3, ]

## Connect to putative enhanceosomes
beta_enhancer_regions <-
  get_regions_of_interest(
    putative_enhanceosome_fantom,
    beta.enriched)

head(as.data.frame(
  methods::slot(beta_enhancer_regions, "annotations")), 5)
```

These rows are enhanceosome regions that also overlap beta-enriched
differential chromatin accessibility calls, combining the TF
co-binding signal with cell-type-specific accessibility evidence.
One such region is rendered below.

```{r fig-beta-track, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_tracks(
  beta_enhancer_regions,
  index = which(names(
    methods::slot(beta_enhancer_regions, "annotations")) == "305"),
  database = database,
  track_connection = track_connection)
```

# Gene-centered visualization

`plot_gene_tracks()` provides a gene-focused view combining
all database tracks (histone marks, ChIP, annotations) with BigWig
signal---no enhanceosome pipeline required.

```{r fig-gene-ins, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_gene_tracks("INS",
  database,
  track_connection)
```

The *INS* view centres on the beta-cell insulin locus so the ATAC
and RNA tracks peak sharply relative to histone-mark context.

```{r fig-gene-gcg, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_gene_tracks("GCG",
  database,
  track_connection)
```

The *GCG* view shows the alpha-cell glucagon locus for comparison;
the signal pattern should mirror-image *INS*, with alpha-specific
accessibility and transcription dominating.

```{r fig-gene-pdx1, fig.width = 14, fig.height = 12, eval = has_bigwigs}
## Wider padding captures distal regulatory elements
plot_gene_tracks("PDX1",
  database,
  track_connection,
  padding = 5000)
```

The *PDX1* view uses wider padding (`padding = 5000`) to expose
distal regulatory elements around this master transcription factor.

# Chromatin state classification

`classify_chromatin_states()` classifies genomic regions by their
histone mark combinations into six biologically meaningful states:
active, bivalent, poised, primed, repressed, and unmarked. Classification
follows ChromHMM/Roadmap Epigenomics conventions (Ernst & Kellis 2012;
Kundaje et al. 2015), with TSS proximity refinement.


```{r chromatin-states, eval = has_data}
chromatin_states <-
  classify_chromatin_states(database)

table(chromatin_states$chromatin_state)
table(chromatin_states$genomic_context)


pdx1_states <- chromatin_states[
  chromatin_states$seqnames == "chr13" &
    chromatin_states$start >= 27900000 &
    chromatin_states$end   <= 28000000, ]
head(pdx1_states, 10)
```

### Visualize the PDX1 locus with chromatin state tracks overlaid:

```{r fig-chromatin-pdx1, fig.width = 14, fig.height = 12, eval = has_bigwigs}
plot_gene_tracks("PDX1",
  database,
  track_connection,
  chromatin_states = chromatin_states,
  show_chromatin = TRUE,
  show_enhancer_highlight = TRUE,
  padding = 5000)
```


# Session Information

```{r sessioninfo}
sessionInfo()
```
